Methods for Analyzing Models of\\ Rod-Shaped Bacteria

### 2025-09-05 paper/main.tex
switch order of setions

%###################################################################################################
\section{Methodology}
% --------------------------------------------------------------------------------------------------
\subsection{Spatial Representation of Rod-Shaped Bacteria}
\label{subsec:spatial-representation}
This work makes use of a few but powerfull assumptions concerning the mathematical representation of
the cells in question.
In order to describe the spatial dynamics of individual agents, we need to define their position,
shape and how they interact with each other.
In order to provide as much flexibility as possible when constructing new models, we restrict
ourselves to assumptions concerning the position such that researchers can freely experiment with
various shapes and interaction types.
Furthermore, it is very important to strike a good balance between the restrictiveness of our
assumptions and generality in order to provide meaningful results and still be able to apply our
approach to a broader class of models.
Taking into account the natural shape of rod-like bacteria, it is clear that they can not be
represented with a singular point $\vec{x}$ in space.
We could extend this positional definition with a vector $\vec{w}$, indicating the direction of
extension but this would restrict ourselves to fully straight rod-shaped bacteria.
However, multiple experiments have shown that MReB is responsible for the growth and maintenance of
the elongated shape of bacteria~\cite{Erickson2001}.
By means of this protein, rod-shaped bacteria can bend elastically on short time-scales and
deform plastically over longer periods of time.
Our goal is to provide a model which can encapsulate these effects.
Thus, we extend our model from a single point to multiple vertices which are part of a 1-dimensional
polygon with two endpoints $\vec{x}_i$.
In essence, we discretize the (possibly non-straight) rod of the agent.
Figure~\ref{fig:mechanics-bacterium} shows a visualization of this approach.
The angles $\alpha_i$ between vertices $\vec{x}_{i-1},\vec{x}_i$ can be used when modeling the
dynamics of these bacteria.

@@

\paragraph{(C) Cellular}
The key component of this model consists of the cellular representation.

It was long believed that bacteria lack cytoskeletal filaments but with the end of the last century,
it was discovered that the MreB protein fills this role since it “[..] forms an actin-like
cytoskeleton in bacteria [..]”~\cite{Erickson2001}.
It can polymerize and form filaments which behave similar to actin microfilaments~\cite{Dersch2020}.
The effect of MreB on the shape of cells during growth was shown in more detail
by~\cite{Ursell2014}.
They found that MreB localizes specifically at regions of negative curvature and thus introduces
heterogeneous cell growth which results in a bending force during the growth of cells.
It also means that the cells are able to sense the curvature of the cell wall through cytoskeleteal
MreB filaments during growth.
\cite{Bratton2018} showed that the trans-membrane protein RodZ can modulate MreB density and
facilitates local MreB assembly.
This feedback mechanism will result in straight rod-shaped bacteria if not interfered from the
external environment.
These results were also confirmed by the studies of~\cite{Wang2010}.
In their experiments, a single bacterium was fixed at the bottom pole while an optically trapped
bead was placed on the upper pole.
They could now exert a force on the free part and thus study the response of the cell under
different conditions.
When treating cells with A22, an antibiotic which disassembles
MreB~\cite{IWAI2002,Gitai2005,Karczmarek2007,Bean2009}, a higher displacement with identical force
could be measured, thus indicating that MreB increases mechanical rigidity of the cell.

@@

@@

>>>>

@@

\paragraph{(C) Cellular}
It was long believed that bacteria lack cytoskeletal filaments but with the end of the last century,
it was discovered that the MreB protein fills this role since it “[..] forms an actin-like
cytoskeleton in bacteria [..]”~\cite{Erickson2001}.
It can polymerize and form filaments which behave similar to actin microfilaments~\cite{Dersch2020}.
The effect of MreB on the shape of cells during growth was shown in more detail
by~\cite{Ursell2014}.
They found that MreB localizes specifically at regions of negative curvature and thus introduces
heterogeneous cell growth which results in a bending force during the growth of cells.
It also means that the cells are able to sense the curvature of the cell wall through cytoskeleteal
MreB filaments during growth.
\cite{Bratton2018} showed that the trans-membrane protein RodZ can modulate MreB density and
facilitates local MreB assembly.
This feedback mechanism will result in straight rod-shaped bacteria if not interfered from the
external environment.
These results were also confirmed by the studies of~\cite{Wang2010}.
In their experiments, a single bacterium was fixed at the bottom pole while an optically trapped
bead was placed on the upper pole.
They could now exert a force on the free part and thus study the response of the cell under
different conditions.
When treating cells with A22, an antibiotic which disassembles
MreB~\cite{IWAI2002,Gitai2005,Karczmarek2007,Bean2009}, a higher displacement with identical force
could be measured, thus indicating that MreB increases mechanical rigidity of the cell.

@@

% --------------------------------------------------------------------------------------------------
\subsection{Spatial Representation}
\label{subsec:spatial-representation}
This work makes use of a few but powerfull assumptions concerning the mathematical representation of
the cells in question.
In order to describe the spatial dynamics of individual agents, we need to define their position,
shape and how they interact with each other.
In order to provide as much flexibility as possible when constructing new models, we restrict
ourselves to assumptions concerning the position such that researchers can freely experiment with
various shapes and interaction types.
Furthermore, it is very important to strike a good balance between the restrictiveness of our
assumptions and generality in order to provide meaningful results and still be able to apply our
approach to a broader class of models.
Taking into account the natural shape of rod-like bacteria, it is clear that they can not be
represented with a singular point $\vec{x}$ in space.
We could extend this positional definition with a vector $\vec{w}$, indicating the direction of
extension but this would restrict ourselves to fully straight rod-shaped bacteria.
However, multiple experiments have shown that MReB is responsible for the growth and maintenance of
the elongated shape of bacteria~\cite{Erickson2001}.
By means of this protein, rod-shaped bacteria can bend elastically on short time-scales and
deform plastically over longer periods of time.
Our goal is to provide a model which can encapsulate these effects.
Thus, we extend our model from a single point to multiple vertices which are part of a 1-dimensional
polygon with two endpoints $\vec{x}_i$.
In essence, we discretize the (possibly non-straight) rod of the agent.
Figure~\ref{fig:mechanics-bacterium} shows a visualization of this approach.
The angles $\alpha_i$ between vertices $\vec{x}_{i-1},\vec{x}_i$ can be used when modeling the
dynamics of these bacteria.

@@

%###################################################################################################
\section{Methodology}
These sections introduce methods which can be combined in order to estimate the parameters of our
computational model.
It has to be noted that they only depend on the spatial representation described in
section~\ref{subsec:spatial-representation}.
Therefore, they can be applied to a whole class of mechanical models which could make different
choices with respect to the cellular details.
